---
title: "Hierarchical Models"
output: html_document
---

# Hierarchical model 
# Rats experiment
# (BDA3, p. 102)

```{r}
# part of the code below is copied from 
# Aki Vehtari <Aki.Vehtari@aalto.fi>
# Markus Paasiniemi <Markus.Paasiniemi@aalto.fi>

library(ggplot2)
library(gridExtra)
library(tidyr)

y <- c(0,0,0,0,0,0,0,0,0,0,0,0,0,0,1,1,1,1,1,1,1,1,2,2,2,2,2,2,2,2,
       2,1,5,2,5,3,2,7,7,3,3,2,9,10,4,4,4,4,4,4,4,10,4,4,4,5,11,12,
       5,5,6,5,6,6,6,6,16,15,15,9,4)
n <- c(20,20,20,20,20,20,20,19,19,19,19,18,18,17,20,20,20,20,19,19,18,18,25,24,
       23,20,20,20,20,20,20,10,49,19,46,27,17,49,47,20,20,13,48,50,20,20,20,20,
       20,20,20,48,19,19,19,22,46,49,20,20,23,19,22,20,20,20,52,46,47,24,14)

x <- seq(0.0001, 0.9999, length.out = 1000)
summary(y/n)
hist(y/n)
```

## Fit beta-binomial with flat prior separately 
## Each experiment has a different theta

```{r}

# helperf function to evaluate density over observations
bdens <- function(n, y, x)
  dbeta(x, y+1, n-y+1)

df_sep <- mapply(bdens, n, y, MoreArgs = list(x = x)) %>%
  as.data.frame() %>% cbind(x) %>% gather(ind, p, -x)

labs1 <- paste('posterior of', c('theta_j', 'theta_71'))

# plot the separate model
plot_sep <- ggplot(data = df_sep) +
  geom_line(aes(x = x, y = p, color = (ind=='V71'), group = ind)) +
  labs(x = expression(theta), y = '', title = 'Separate model', color = '') +
  scale_y_continuous(breaks = NULL) +
  scale_color_manual(values = c('blue','red'), labels = labs1) +
  theme(legend.background = element_blank(), legend.position = c(0.8,0.9))
# The last one is for emphasize colored red
plot_sep
```

## Fit pooled Beta-Binomial model (common theta)

```{r}

df_pool <- data.frame(x = x, p = dbeta(x, sum(y)+1, sum(n)-sum(y)+1))

plot_pool <- ggplot(data = df_pool) +
  geom_line(aes(x = x, y = p, color = '1')) +
  labs(x = expression(theta), y = '', title = 'Pooled model', color = '') +
  scale_y_continuous(breaks = NULL) +
  scale_color_manual(values = 'red', labels = 'Posterior of common theta') +
  theme(legend.background = element_blank(), legend.position = c(0.7,0.9))

# Plot both separate and pooled model
grid.arrange(plot_sep, plot_pool)
```

## Hierarchical model (Beta-Binomial)

```{r}
A <- seq(0.1, 10, length.out = 200) ## alpha
B <- seq(3, 60, length.out = 200) ## beta
# make vectors that contain all pairwise combinations of A and B
cA <- rep(A, each = length(B))
cB <- rep(B, length(A))
# Use logarithms for numerical accuracy!
lpfun <- function(a, b, y, n) 
  sum(lgamma(a+b)-lgamma(a)-lgamma(b)+lgamma(a+y)+lgamma(b+n-y)-lgamma(a+b+n))

lp <- mapply(lpfun, cA, cB, MoreArgs = list(y, n))
df_marg <- data.frame(x = cA, y = cB, p = exp(lp - max(lp)))

# Subtract maximum value to avoid over/underflow in exponentation
title1 <- 'Contour of likelihood for alpha beta'
# create a plot of the marginal posterior density
postdensityalphabeta = ggplot(data = df_marg, aes(x = x, y = y)) +
  geom_raster(aes(fill = p, alpha = p), interpolate = T) +
  geom_contour(aes(z = p), colour = 'black', size = 0.2) +
  coord_cartesian(xlim = range(cA), ylim = range(cB)) +
  labs(x = 'alpha', y = 'beta', title = title1) +
  scale_fill_gradient(low = 'yellow', high = 'red', guide = F) +
  scale_alpha(range = c(0, 1), guide = F)
postdensityalphabeta
```


## First visualize(calculate) the marginal posterior distribution of alpha and beta after integrating out theta's
```{r}

# Compute the marginal posterior of alpha and beta in hierarchical model
# Use grid
A <- seq(0.5, 6, length.out = 100) ## alpha
B <- seq(3, 33, length.out = 100) ## beta
# make vectors that contain all pairwise combinations of A and B
cA <- rep(A, each = length(B))
cB <- rep(B, length(A))

# Use logarithms for numerical accuracy!
lpfun <- function(a, b, y, n) log(a+b)*(-5/2) +
  sum(lgamma(a+b)-lgamma(a)-lgamma(b)+lgamma(a+y)+lgamma(b+n-y)-lgamma(a+b+n))

lp <- mapply(lpfun, cA, cB, MoreArgs = list(y, n))
df_marg <- data.frame(x = cA, y = cB, p = exp(lp - max(lp)))

# Subtract maximum value to avoid over/underflow in exponentation
title1 <- 'The marginal posterior of alpha and beta in hierarchical model'
# create a plot of the marginal posterior density
postdensityalphabeta = ggplot(data = df_marg, aes(x = x, y = y)) +
  geom_raster(aes(fill = p, alpha = p), interpolate = T) +
  geom_contour(aes(z = p), colour = 'black', size = 0.2) +
  coord_cartesian(xlim = c(1,5), ylim = c(4, 26)) +
  labs(x = 'alpha', y = 'beta', title = title1) +
  scale_fill_gradient(low = 'yellow', high = 'red', guide = F) +
  scale_alpha(range = c(0, 1), guide = F)
postdensityalphabeta
```


## obtain random samples from p(alpha,beta|data)
## (samp_A, samp_B) contains 100 samples of alpha and beta
```{r}
# df_marg: first column (value of alpha)
#          second column (value of beta)
#          third column (value of p(alpha, beta | data))
# sample from the grid (with replacement)
nsamp <- 100
samp_indices <- sample(length(df_marg$p), size = nsamp,
                       replace = T, prob = df_marg$p/sum(df_marg$p))

samp_A <- cA[samp_indices[1:nsamp]]
samp_B <- cB[samp_indices[1:nsamp]]
```

## visualize the samples
```{r}
nsamp2 <- 2000
samp_indices2 <- sample(length(df_marg$p), size = nsamp2,
                       replace = T, prob = df_marg$p/sum(df_marg$p))
samplesalphabeta = data.frame(alpha = cA[samp_indices2[1:nsamp2]], beta = cB[samp_indices2[1:nsamp2]])
scatteralphabeta = ggplot(samplesalphabeta, aes(x=alpha, y=beta)) + geom_point()
grid.arrange(scatteralphabeta, postdensityalphabeta)
```

## transformation
```{r}
samplesalphabetatransform = samplesalphabeta
samplesalphabetatransform$alpha = log(samplesalphabeta$alpha/samplesalphabeta$beta)
samplesalphabetatransform$beta = log(samplesalphabeta$alpha+samplesalphabeta$beta)
scatteralphabetatransform = ggplot(samplesalphabetatransform, aes(x=alpha, y=beta)) + geom_point() + labs(x = "log(alpha/beta)", y = "log(alpha+beta)")
print(summary(samplesalphabetatransform$alpha))
print(summary(samplesalphabetatransform$beta))
scatteralphabetatransform
```


## calculate beta density at samples of p(alpha,beta|data)

```{r}
df_psamp <- mapply(function(a, b, x) dbeta(x, a, b),
                   samp_A, samp_B, MoreArgs = list(x = x)) %>%
  as.data.frame() %>% cbind(x) %>% gather(ind, p, -x)

# helper function to convert ind to numeric for subsetting
indtonum <- function(x) strtoi(substring(x,2))

# Plot samples from the distribution of distributions Beta(alpha,beta),
# that is, plot Beta(alpha,beta) using posterior samples of alpha and beta
title2 <- 'Posterior samples from the distribution of distributions Beta(a,b)'
plot_psamp <- ggplot(data = subset(df_psamp, indtonum(ind) <= 20)) +
  geom_line(aes(x = x, y = p, group = ind)) +
  labs(x = expression(theta), y = '', title = title2) +
  scale_y_continuous(breaks = NULL)
plot_psamp
```

## calculate predictive distribution for new theta

```{r}

# The average of above distributions, is the predictive distribution
# for a new theta, and also the prior distribution for theta_j
df_psampmean <- spread(df_psamp, ind, p) %>% subset(select = -x) %>%
  rowMeans() %>% data.frame(x = x, p = .)

title3 <- 'Predictive distribution for a new theta and prior for theta_j'
plot_psampmean <- ggplot(data = df_psampmean) +
  geom_line(aes(x = x, y = p)) +
  labs(x = expression(theta), y = '', title = title3) +
  scale_y_continuous(breaks = NULL)
plot_psampmean
```



```{r}

# combine the plots
grid.arrange(plot_psamp, plot_psampmean)

```

## compare separate model and hierarchical model

```{r}

# And finally compare the separate model and hierarchical model
# (using every seventh sample for clarity)

# the separate model
plot_sep7 <- ggplot(data = subset(df_sep, indtonum(ind)%%7==0)) +
  geom_line(aes(x = x, y = p, color = (ind=='V49'), group = ind)) +
  labs(x = expression(theta), y = '', title = 'Separate model', color = '') +
  scale_y_continuous(breaks = NULL) +
  scale_color_manual(values = c('blue', 'red'), guide = F) +
  theme(legend.background = element_blank(), legend.position = c(0.8,0.9))

# the hierarchical model
# Note that these marginal posteriors for theta_j are more narrow than
# in the separate model case, due to the borrowed information from the
# other theta_j's

# average density over samples (of a and b) for each (n,y)-pair
# at each point x
bdens2 <- function(n, y, a, b, x)
  rowMeans(mapply(dbeta, a + y, n - y + b, MoreArgs = list(x = x)))

df_hier <- mapply(bdens2, n, y, MoreArgs = list(samp_A, samp_B, x)) %>%
  as.data.frame() %>% cbind(x) %>% gather(ind, p, -x)

plot_hier7 <- ggplot(data = subset(df_hier, indtonum(ind)%%7==0)) +
  geom_line(aes(x = x, y = p, color = (ind=='V49'), group = ind)) +
  labs(x = expression(theta), y = '', title = 'Hierarchical model', color = '') +
  scale_color_manual(values = c('blue', 'red'), guide = F) +
  scale_y_continuous(breaks = NULL) +
  theme(legend.background = element_blank(), legend.position = c(0.8,0.9))

grid.arrange(plot_sep7, plot_hier7)

```

## use posterior samples of theta to approximate 
## closed form (beta) posterior distributions of theta

```{r}
# sample from posterior distribution of theta_j

nsamp <- 10000
samp_indices <- sample(length(df_marg$p), size = nsamp,
                       replace = T, prob = df_marg$p/sum(df_marg$p))

samp_A <- cA[samp_indices[1:nsamp]]
samp_B <- cB[samp_indices[1:nsamp]]
samplestheta <- matrix(0, nsamp, length(y))
for(j in 1:length(y)){
  samplestheta[, j] = sapply(1:nsamp, function(k) rbeta(1, samp_A[k]+y[j], samp_B[k]+n[j]-y[j]))
}
breakpoints <- seq(0, 1, length.out = 200)
xx <- hist(samplestheta[, 1], breaks = breakpoints, plot = FALSE)$mids
densitytheta <- matrix(0, length(breakpoints) - 1, length(y))
for(j in 1:length(y)){
  densitytheta[, j] <- hist(samplestheta[, j], breaks = breakpoints, plot = FALSE)$density
}
df_hier_samp <-densitytheta %>%
  as.data.frame() %>% cbind(xx) %>% gather(ind, p, -xx)
plot_hier7_samp <- ggplot(data = subset(df_hier_samp, indtonum(ind)%%7==0)) +
  geom_line(aes(x = xx, y = p, color = (ind=='V49'), group = ind)) +
  labs(x = expression(theta), y = '', title = 'Hierarchical model', color = '') +
  scale_color_manual(values = c('blue', 'red'), guide = F) +
  scale_y_continuous(breaks = NULL) +
  theme(legend.background = element_blank(), legend.position = c(0.8,0.9))
grid.arrange(plot_hier7_samp, plot_hier7)

```

## making inference on theta's

```{r}
nsamp <- 1000
samp_indices <- sample(length(df_marg$p), size = nsamp,
                       replace = T, prob = df_marg$p/sum(df_marg$p))

samp_A <- cA[samp_indices[1:nsamp]]
samp_B <- cB[samp_indices[1:nsamp]]
samplestheta <- matrix(0, nsamp, length(y))
for(j in 1:length(y)){
  samplestheta[, j] = sapply(1:nsamp, function(k) rbeta(1, samp_A[k]+y[j], samp_B[k]+n[j]-y[j]))
}
postintervals_sample <- apply(samplestheta, 2, function(x) c(median(x), c(quantile(x, c(0.025, 0.975)))))

```

## visualize results and compare

```{r}

rawest <- jitter(y / n)
plot(rawest, postintervals_sample[1, ], pch = 19, col = 'red', cex = 0.5, ylim = c(-0.01, 0.4), ylab = 'posterior intervals', xlab = 'raw estimator')
for(k in 1:length(y)){
  lines(cbind(rep(rawest[k], 2), postintervals_sample[2:3, k]))
}
lines(seq(0, 0.4, by = 0.01), seq(0, 0.4, by = 0.01), col = 'blue')
phatpool = sum(y)/sum(n)
points(phatpool, phatpool, cex = 2, col = 'green')
lines(cbind(rep(qbeta(0.5, sum(y)+1, sum(n)-sum(y)+1), 2), c(qbeta(0.025, sum(y)+1, sum(n)-sum(y)+1), qbeta(0.975, sum(y)+1, sum(n)-sum(y)+1))), col = 'yellow', lwd = 4)
```





